Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-28 supplement.tex
Add alternative strategy and description for complexity reduction due to bipartiteness (and constraints in more generality)

\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively. It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}}
To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp. \Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.

\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
    the specialization of constraint networks to RNA design
    networks. Luckily, there should be an even easier way to describe
    the complexity reduction without introducing new variables; ideally let's discuss}}

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.

>>>>

\section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}

\textcolor{red}{\textbf{SW: Simpler alternative to
    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
    with the (recent) specialization of constraint networks to RNA design
    networks.}}

While Alg.~\ref{alg:pf} computes messages values for
\emph{all} possible combinations of nucleotides for the positions in
the separator set, we observe here that this is \emph{not} required for computing all relevant
partition functions. In particular, the algorithm can be restricted to
consider only \emph{valid} combinations, which satisfy the (hard)
constraints on the nucleotides.

RNA design introduces constraints due to the requirement of canonical
(aka complemenentary) base pairing, which can be exploited in a
particularly simple and effective way.  As already noted by
\citet{Flamm2001}, the base pair complementarity induces a
bi-partition of each connected component within the base pair
dependency graph, such that the nucleotides assigned to the two set of
nodes in the partition are restricted to values $\{\Ab,\Gb\}$ and
$\{\Cb,\Ub\}$ respectively; we call a partitial sequence
\emph{cc-valid}, iff its determined positions satisfy this separation
for all determined positions of the same connected component.

% Let $CC=\{cc_1$, $cc_2, \dots\}$ denote the vertex sets of the
% connected components of the base pair dependency graph. In this graph,
% let each component have the bi-partition $cc_i=cc^0_i\cup cc^1_i$.

One can now modify Alg.~\ref{alg:pf}, on tree decomposition
$(T,\chi)$, such that the message values
$\evalfor{\Message{u}{v}}{\val}$ are computed only for cc-valid
partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the
loop over $\val'\in\partseqs(\difference{u}{v})$ is restricted, such
that $\substitute{\val}{\val'}$ are cc-valid. (The analogous change is
made to the sampling algorithm Alg.~\ref{alg:sampling}.)

The correctness of the modified algorithm still follows from an
induction argument: in this way, the message computation of one node
evaluates messages from its children only at cc-valid partial
sequences. The result of this computation is a message, which is
defined at all cc-valid partial sequences.

Remarkably, this strongly improves the algorithm's time
complexity. Where the original algorithm enumerates all partial
sequences $\substitute{\val}{\val'}$, where
$\val\in\partseqs(\separator{u}{v})$ and
$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
enumerates in two stages:
\begin{enumerate}
\item for each connected component of the base pair dependency graph
  that involves any position in
  $\separator{u}{v}\cup\difference{u}{v} = \chi(v)$, it first
  enumerates all possibilities how the positions from the partition
  sets are assigned to either $\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$
\item it enumerates the possible partial sequences from the (according
  to the decisions of the first stage) restricted domains of the
  position in $\chi(v)$.
\end{enumerate}
Let $c(v)$ be the number of connected components involving positions
in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
combinations. For each such combination, the second stage enumerates
only $2^{|\chi(v)|}$ partial sequences, since there remain only two
possible nucleotide choices for each position. All message values for
node $v$ are thus computed in time
$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
length, $k$ is the number of structures, and
$\hat w := \max_{v\in T} c(v)+|\chi(v)|$. Notably, $\hat w$ can not
exceed $2(w+1)$ such that the modified algorithm does not increase the
original complexity.
%
As useful corollary, one obtains the (looser) complexity
bound $$O(nk2^{w+1}2^{cc})$$ in terms of treewidth $w$ and connected
components $cc$.

Finally, we remark that even stronger time savings could be possible
in practice, since cc-valid partial sequences can still violate
complementarity constraints, e.g. by assigning C and A to positions in
different sets of a partition, thus satisfying the bi-partition
constraints, where the positions base pair directly, rendering the
partial sequence invalid. Moreover, applications of the sampling
framework, can introduce additional constraints that further reduce
the number of valid partial subsequences. However, exploiting all such
constraints, in a complete and general way, likely causes significant
overhead in an implementation and does not straightforwardly improve
the theoretical bound.

\section{Exploiting the bipartite nature of base pair dependency
  graphs to reduce the complexity}\label{sec:first-improvedComplexity}
\textcolor{red}{\textbf{TODO: discuss and merge with/replace by previous Section}}

As already noted by \citet{Flamm2001}, the requirement of canonical
base pairing induces a bi-partition of each connected component within
the base pair dependency graph, such that the nucleotides assigned to
the two set of nodes in the partition are restricted to values
$\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
%
It follows that, within each connected component of the base pair
dependency graph, the assignment of any position drastically restricts
the domain of the remaining positions, \emph{i.e.} they may now be
assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}} To exploit this property, and further
reduce the exponential factor of the complexity, we introduce the
notion of \Def{parity} of a vertex/position, whereby vertices of
\Def{even} (resp. \Def{odd}) polarities may only be assigned values in
$\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must
be even and vice versa, so the choice of a parity for any single
vertex caracterizes the parity of all vertices in a bipartite
connected subgraph.

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the base pair dependency graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the base pair dependency graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
